---
title: "NYPDShootingAnalysisProject"
author: "Maris"
date: "2023-09-19"
output:
  html_document: default
  pdf_document: default
---

#### Libraries used
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(ggplot2)
library(vcd)
```

## NYPD Shooting Incident Data (Historic)

####
List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity.

### Getting the data
```{r get_shooting_data}
# getting the link first
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# reading in the data from the link
nypd_ytd <- read.csv(url_in)
```

### Areas of analysis

I wanted to look into three aspects of the data.

1. Shooting location vs time of day (nighttime or daytime).

2. Statistical murder vs the time of day.

3. Shooting frequency distribution for days of the week



### Removing unnecessary columns
```{r remove_columns}
# removing various columns
nypd_ytd <- nypd_ytd %>% 
  select(
    -Latitude,
    -Longitude,
    -Lon_Lat,
    -X_COORD_CD,
    -Y_COORD_CD,
    -PERP_RACE,
    -VIC_RACE,
    -PERP_AGE_GROUP,
    #-OCCUR_TIME,
    -BORO,
    -LOC_OF_OCCUR_DESC,
    -JURISDICTION_CODE,
    -LOCATION_DESC,
    -PRECINCT,
    #-STATISTICAL_MURDER_FLAG,
    -PERP_SEX,
    -VIC_SEX,
    -VIC_AGE_GROUP
    )
```

## 1 Shooting location vs daylight or nighttime hours

### Converting date strings to date and time objects
```{r convert_date_time}
# converting date strings to date objects
nypd_ytd <- nypd_ytd %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))

# converting time strings (characters) to time objects (this is a Period class object)
nypd_ytd <- nypd_ytd %>% mutate(OCCUR_TIME = hms(OCCUR_TIME))
```

### Let's add a rough estimate of daytime/nightime for each shooting incident

Here I am simply defining the DAYLIGHT as being between 5am and 10pm. It is not ideal
because sunset and sunrise times vary during the year but this a "one size fits all"
approach because I couldn't find sunset/sunrise data for all the years the data covers.
```{r add_daylight}
# adding a column that will contain True of False for daylight
# we'll define the daylight as True between 5am and 10pm
nypd_ytd <- nypd_ytd %>%
  mutate(
    DAYLIGHT = (OCCUR_TIME >= hms("05:00:00") & OCCUR_TIME <= hms("22:00:00") )
    )
```

### Let's add a column for day of the week for each shooting
```{r add_day}
# adding a column that will contain day of the week
nypd_ytd$OCCUR_WEEKDAY <- weekdays(nypd_ytd$OCCUR_DATE)
```

### Let's factor the days of the week, location, and stat murder flag classifications
```{r factor_days_location}
nypd_ytd$OCCUR_WEEKDAY <- as.factor(nypd_ytd$OCCUR_WEEKDAY)
nypd_ytd$LOC_CLASSFCTN_DESC <- as.factor(nypd_ytd$LOC_CLASSFCTN_DESC)
nypd_ytd$STATISTICAL_MURDER_FLAG <- as.factor(nypd_ytd$STATISTICAL_MURDER_FLAG)
```

### Removing NA and blanks from location classification

A large number of the location classification entries were blank, so I removed them.
```{r removing_na}
nypd_subset <- nypd_ytd[!(nypd_ytd$LOC_CLASSFCTN_DESC == "" |
                          is.na(nypd_ytd$LOC_CLASSFCTN_DESC)), ]

```

```{r summary_post}
subset_summary = summary(nypd_subset)
print(subset_summary)
```

### Using the subset data to generate a mosaic plot using mosaic function
Y axis labels are overlapping the graph which I unfortunately wasn't able to fix.
```{r location_daylight}
mosaic(~ LOC_CLASSFCTN_DESC + DAYLIGHT, data = nypd_subset,
       split_vertical = FALSE, # we want the rectangles to vary in width not height
       # ensuring labels are horizontal and smaller font so they don't overlap
       labeling_args = list(rot_labels = c(0, 0), gp_labels = gpar(fontsize = 6)) 
       )
```

#### Same location classification and daylight data in number form

We can see that during our defined daylight there are considerably more shootings 
happening indoors - DWELLING, HOUSING. Transit shootings also seem to be much more
prevalent during the daylight hours.

```{r contingency_table_1}
contingency_table <- table(nypd_subset$LOC_CLASSFCTN_DESC, nypd_subset$DAYLIGHT)
print (contingency_table)
```







## 2 Statistical murder flag vs daylight or nighttime hours

### Let's see if the daylight and statistical murder flag factors are independent
Using the Chi-Square test to test the independence of statistical murder flag and
daylight
```{r independece_test}
# checking for independence of two factors
chisq_test_loc <- chisq.test(table(nypd_ytd$DAYLIGHT, nypd_ytd$STATISTICAL_MURDER_FLAG))

chisq_test_loc
```

#### It's resulting in a low p-value, let's see what the plot looks like.

### Let's try to plot the DAYLIGHT and STATISTICAL_MURDER_FLAG using a mosaicplot
```{r stat_murder_flag_vs_daylight}
relationship_table <- table(nypd_ytd$DAYLIGHT, nypd_ytd$STATISTICAL_MURDER_FLAG)

mosaicplot(relationship_table, 
           main="Mosaic Plot of Daylight and Statistical Murder Flag", 
           xlab="Daylight", 
           ylab="Statistical Murder Flag",
           las = 1 #making labels appear horizontally
           )
```

### Daytime vs nighttime lethality of shootings.

Shootings that take place during our defined "daytime" hours seem to be slightly
more likely to result in death. Potentially, the better visibility during the 
daytime hours could be a factor that improves shooting accuracy, hence the 
outcomes are more tragic. However, this could simply be noise.

```{r contingency_table_2}
contingency_table <- table(nypd_ytd$STATISTICAL_MURDER_FLAG, nypd_ytd$DAYLIGHT)
print (contingency_table)
```




## 3 Shooting frequency distribution per weekday

Let's look at the frequency distribution per weekdays to see if some days are 
more "popular" than others.
```{r bar_plot}
# first ordering the weekdays in correct order
nypd_ytd$OCCUR_WEEKDAY <- factor(nypd_ytd$OCCUR_WEEKDAY, 
                                 levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# putting the frequency distribution into tab variable
observed <- table(nypd_ytd$OCCUR_WEEKDAY)

# if random, we would expect a uniform distribution
expected <- rep(sum(observed)/length(observed), length(observed))


# first we'll shorten the names of the weekdays
original_names <- levels(factor(nypd_ytd$OCCUR_WEEKDAY))
short_names_map <- c(
                        "Monday" = "Mon",
                        "Tuesday" = "Tue",
                        "Wednesday" = "Wed",
                        "Thursday" = "Thu",
                        "Friday" = "Fri",
                        "Saturday" = "Sat",
                        "Sunday" = "Sun"
                        )
short_names <- short_names_map[original_names]

barplot(observed,
        main = "Shooting frequency distribution by weekdays",
        ylim = c(0, max(c(observed, expected)) + 5), 
        xlab = "Weekday", 
        ylab = "Frequency", 
        col = rgb(red = 0.53, green = 0.81, blue = 0.92, alpha = 0.5),
        # let's shorten the weekday names
        names.arg = short_names,
)
```

Sundays and Saturdays are certainly more "popular" than the rest of the week.

### Checking if the distribution is random
```{r randomness_check}
# Chi-square goodness of fit test
chisq_test_dist <- chisq.test(observed, p = rep(1/length(levels(nypd_ytd$OCCUR_WEEKDAY)), length(levels(nypd_ytd$OCCUR_WEEKDAY))))

# let's see the result

print(chisq_test_dist)
```

The p-value is well below 0.05 so there is evidence to support the rejection of 
the null hypothesis (uniform distribution) - some days seem to be more 
prevalent than others.


## Conclusion
I examined three aspects of the NYPD shooting incident data (Year to date).

1) The first aspect was to see whether there is any relationship between the 
location classification of the shooting and the time of day (daylight hours or 
nighttime hours). I discovered that there are some differences between the of
day when the shooting takes place and the location that it occurs. For example, 
street shootings and transit shootings are more prevalent during the daytime hours 
and shootings with a vehicle involved are more prevalent during the nighttime hours.

2) The second aspect - looking at shootings resulting in death, there seems to
be a slightly higher chance of dying if getting shot during the day, however, the
difference is marginal and it could simply be due to noise.

3) The third aspect - looking at the weekday distribution of shootings, it is
clear that Saturdays and Sundays have more shootings take place with Sunday
coming out on top. However, I was surprised to see that Mondays are beating 
Fridays in their shooting count, I would have expected Fridays to see more 
action than Mondays.

#### Possible sources of bias
The data on the first aspect wasn't great because a considerable number of entries
had no location classification so a lot of rows were excluded from the analysis.
If the information was there, it may have changed the result.

On the second aspect, the effect was marginal, so this could simply be noise.

On the third aspect, the result was close to what I would expect.

Also, significantly, for the first and second aspect I classified "daytime" as being 
between 5am and 10pm which is not ideal. Ideally, it would be better to have a sliding
daytime/nighttime value that corresponds to astronomical twilight time for the
specific day of the year. This would result in a more accurate analysis.

Another possible issue could be that have made some technical/statistical
mistakes in my analysis by using the wrong approach or incorrect filtering, etc.,
which has resulted in erroneous results.



```{r session}
sessionInfo()
```
